---
title: "Impact of Severe Weather on Population Health and Economy in the US"
author: "Ayo Johnson"
date: "5/9/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction

This project explores data from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. Its aim is to determine:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

Policy makers may use its findings as a guide for resource allocation before during and after severe weather events.

## Data Processing

Libraries used for this project
```{r}
library(data.table)
library(lubridate)
library(date)
library(dtplyr)
library(ggplot2)
```

Loading Data and Preprocessing

```{r}
knitr::opts_chunk$set(cache = TRUE)
StormData <- fread ("/Users/ayojohnson/Coursera/ReproducibleResearch/repdata-data-StormData.csv")
```
```{r}
dim(StormData)
head(StormData, n = 2)
tail(StormData, n = 2)
```

Checking where the NAs are
```{r}
colSums(is.na(StormData))
```

Checking column names

```{r}
colnames(StormData)
```
Checking data structure
```{r}
str(StormData)
```


checking again
```{r}
tables()
```
Verifying that there are 48 Event types
```{r, checkunique}
Events <- unique(StormData$EVTYPE)
length(Events)
```

For the purpose of analysing the impact of different weather events on the economy and public health across the United States the following variables are most relevant:

 - EVTYPE : type of weather event
 - STATE : State
 - FATALITIES : Number of deaths
 - INJURIES : Number of injuries
 - PROPDMG : Value of property damage
 - CROPDMG : Value of crop damage
 - PROPDMGEXP : exponent of value of property damage
 - CROPDMGEXP: exponent of value of cost damage

According to the official documentation of the data, collection of storm events began in 1950 but it was only until 1996 that the full spectrum of weather events began to be collected.
The documentation also notes that each entry is given a reference number (variable Refnum). It appears from our initial examination that the refnum are in numerical order.
For these reasons we will retain the variables REFNUM and BGN_DATE (date of start of event). This enables random sampling of the data as well as analysis of weather events by year. 
Including the state allows analysis by state - assuming, for instance, that some states may have characteristics - such as architecture and geography which might affect impacts of severe storms.

With these points in mind we'll subset the data to include only the variables identified above from the year 1996.

Now we can select the relevant data
```{r}
StormData <- select(StormData,REFNUM,EVTYPE,BGN_DATE,STATE,FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP)
head (StormData, n = 2)
##table(StormData)
tables()
str(StormData)
```
House-keeping: Convert BGN_DATE to date format

```{r}
#StormData$BGN_DATE <- as.Date(StormData$BGN_DATE, "%m/%d/%Y %H:%M:%S")
StormData$BGN_DATE <- mdy_hms(StormData$BGN_DATE)
class(StormData$BGN_DATE)
dim(StormData)
str(StormData)
```


Now we select only those records with a BGN_DATE post 01/01/1996
```{r}
StormData <- filter(StormData, year(BGN_DATE) >= "1996")
head(StormData)
tail(StormData)
#tables(StormData)
dim(StormData)
```

Let's take an exploratory look at this reduced dataset focusing on Storm events variable EVTYPE
change StormData$BGN_DATE to numeric first

```{r}
##StormData$BGN_DATE <- as.numeric(StormData$BGN_DATE)
```

creating a Year column
```{r}
StormData$Year <- year(StormData$BGN_DATE)
str(StormData)
```

```{r}
ggplot(data = StormData, aes(x = Year))+
        geom_histogram(binwidth = 1,
                       col = "red",
                       fill = "yellow",
                       alpha = .2)+
        labs(title = "Histogram for Year")+
        labs(x = "Year", y = "Number of events")+
        scale_x_continuous(breaks = seq(1996, 2011, 1))+
        scale_y_continuous(breaks=seq(0,65000,5000))
```

Checking that there are 48 different weather event types as noted in the code documentation.

```{r}
Events <- unique(StormData$EVTYPE)
length(Events)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
